---
title: "MAST90138 Asmt1"
author: "Kechen Zhao 957398"
date: "15/08/2021"
output: word_document
---

Problem 1

a. To make $$ \Sigma $$ to be the covariance matrix, it needs to be symmetric and positive semi-definite
$$ \Sigma  = \begin{bmatrix} 1&2 \\ a&b \\ \end{bmatrix}$$
To make it to be symmetric, let a = 2, then find the value of b to make the matrix positive semi-definite (can check whether it is positive semi-definite or not by checking whether if all its eigenvalues are non-negative):
Let A = $$ \Sigma $$, then the eigenvalue problem becomes $$ A*v = \lambda*v => A*v - \lambda*v = 0 => (A-\lambda * I)*v = 0$$
If all eigenvectors v are non-zero, then the solution will be given by $$ A-\lambda * I  = 0 $$
In matrix form, $$ |A-\lambda * I| = |\begin{bmatrix} 1&2 \\ 2&b \\ \end{bmatrix} - \begin{bmatrix} \lambda&0 \\ 0&\lambda \\ \end{bmatrix}| = |\begin{bmatrix} 1-\lambda&2 \\ 2&b-\lambda \\ \end{bmatrix}| = 0$$
$$ \lambda^2-(b+1)\lambda+(b-4) = 0$$
We will then find the value of b which makes all solutions of lambda non-negative:
Case 1. lambda = 0, then $$ 0-(b+1)*0 + b-4 = 0 => b = 4$$, the $$ \Sigma  = \begin{bmatrix} 1&2 \\ 2&4 \\ \end{bmatrix}$$, calculate its eigenvalues to check positive semi-definite:
```{r}
sigma = matrix(c(1,2,2,4), ncol=2, nrow=2)
EE = eigen(sigma)
Evalues = EE$values
Evalues
```
Eigenvalues = 5 and 0 which are all non-negative, so b=4 meets the requirement.
Case 2. lambda > 0, then based on the characteristic of quadratic equation, roots are real and positive if and only if $$b+1 > 0$$, $$(b+1)^2-4*(b-4) > 0$$ and $$4*(b-4) >0$$. Combining the solutions of all conditions, lambda > 0 if and only if b > 4.
Checking the roots when b > 4: 
The solutions of $$ \lambda^2-(b+1)\lambda+(b-4) = 0$$ is $$ \lambda_1 = \frac{(b+1)-\sqrt{(b+1)^2-4*(b-4)}}{2},  \lambda_2 = \frac{(b+1)+\sqrt{(b+1)^2-4*(b-4)}}{2}$$ 
By ploting the equation $$ (b+1)-\sqrt{(b+1)^2-4*(b-4)} $$, we can see its a monotoinc increasing funnction with value 0 when b = 4, which proves that values of lambda will always be real and positive if b > 4. 
```{r}
eq = function(b){(b+1)-sqrt((b+1)^2-4*(b-4))}
plot(eq(1:100), type='l')
```
Thus, combining the results of both cases, values of lambda will be non-negative if and only if b >= 4, which makes the matrix sigma to be positive semi-definite. Therefore, sigma can be a covariate matrix when a = 2 and b >= 4. 

b. $$ \Sigma  = \begin{bmatrix} 13&-4 \\ -4&7 \\ \end{bmatrix}$$, the eigenvalue problem is $$ \Sigma*v = \lambda*v => \Sigma*v - \lambda*v = 0 => (\Sigma-\lambda * I)*v = 0$$
If all eigenvectors v are non-zero, then the solution will be given by $$ A-\lambda * I  = 0 $$
In matrix form, $$ |\Sigma-\lambda * I| = |\begin{bmatrix} 13&-4 \\ -4&7 \\ \end{bmatrix} - \begin{bmatrix} \lambda&0 \\ 0&\lambda \\ \end{bmatrix}| = |\begin{bmatrix} 13-\lambda&-4\\ -4&7-\lambda \\ \end{bmatrix}| = 0$$
This gives rise to the quadratic equationn $$ (13-\lambda)(7-\lambda)-16 = 0 => \lambda^2-20\lambda+75 = 0 => (\lambda-15)(\lambda-5) = 0 => \lambda_1 = 15, \lambda_2 = 5$$
So eigenvalues of matrix sigma is 15 and 5, and their corresponding eigenvectors are:

-for eigenvalue $$ \lambda_1 = 15 $$, the eigenvector can be calculated as $$ (\Sigma-15I)*v_1 = 0 => (\begin{bmatrix} 13&-4 \\ -4&7 \\ \end{bmatrix} - \begin{bmatrix} 15&0 \\ 0&15 \\ \end{bmatrix})*v_1 = \begin{bmatrix} -2&-4\\ -4&-8 \\ \end{bmatrix}* \begin{bmatrix} v_{11}\\ v_{12} \\ \end{bmatrix}= 0$$
This gives $$ -2v_{11}-4v_{12} = 0, -4v_{11}-8v_{12} = 0$$, norm 1 gives another constraints $$ \sqrt{v_{11}^2+v_{12}^2} = 1$$, by solving system of equation we have $$ v_1 = \begin{bmatrix} -2\sqrt{\frac{1}{5}}\\ \sqrt{\frac{1}{5}} \\ \end{bmatrix} ,or, v_1 = \begin{bmatrix} 2\sqrt{\frac{1}{5}}\\ -\sqrt{\frac{1}{5}} \\ \end{bmatrix}$$

-for eigenvalue $$ \lambda_1 = 5 $$, the eigenvector can be calculated as $$ (\Sigma-5I)*v_2 = 0 => (\begin{bmatrix} 13&-4 \\ -4&7 \\ \end{bmatrix} - \begin{bmatrix} 5&0 \\ 0&5 \\ \end{bmatrix})*v_1 = \begin{bmatrix} 8&-4\\ -4&2 \\ \end{bmatrix}* \begin{bmatrix} v_{21}\\ v_{22} \\ \end{bmatrix}= 0$$
This gives $$ 8v_{21}-4v_{22} = 0, -4v_{21}+2v_{22} = 0$$, norm 1 gives another constraints $$ \sqrt{v_{21}^2+v_{22}^2} = 1$$, by solving system of equation we have $$ v_2 = \begin{bmatrix} \sqrt{\frac{1}{5}}\\ 2\sqrt{\frac{1}{5}} \\ \end{bmatrix} ,or, v_2 = \begin{bmatrix} -\sqrt{\frac{1}{5}}\\ -2\sqrt{\frac{1}{5}} \\ \end{bmatrix}$$

So the eigenvector for $$ \lambda_1 = 15$$ is $$ v_1 = \begin{bmatrix} -2\sqrt{\frac{1}{5}}\\ \sqrt{\frac{1}{5}} \\ \end{bmatrix}$$, the eigenvector for $$ \lambda_2 = 5$$ is $$ v_2 = \begin{bmatrix} \sqrt{\frac{1}{5}}\\ 2\sqrt{\frac{1}{5}} \\ \end{bmatrix}$$

Check by using R:
```{r}
sigma = matrix(c(13,-4,-4,7), ncol=2, nrow=2)
EE = eigen(sigma)
Evalues = EE$values
Evalues
Evec = EE$vectors
Evec
```
$$\Gamma = \begin{bmatrix} -2\sqrt{\frac{1}{5}}&\sqrt{\frac{1}{5}} \\ \sqrt{\frac{1}{5}}&2\sqrt{\frac{1}{5}} \\ \end{bmatrix}, \Lambda = \begin{bmatrix} 15&0\\ 0&5 \\ \end{bmatrix}$$
Check by using R:
```{r}
Gamma = matrix(c(-2*sqrt(1/5), sqrt(1/5), sqrt(1/5), 2*sqrt(1/5)), ncol=2, nrow=2)
Lambda = matrix(c(15, 0, 0, 5), ncol=2, nrow=2)
Gamma%*%Lambda%*%t(Gamma)
```

c. 
```{r}
X <- read.table("Wheat data.txt", sep = "")
dim(X)
wheat_variety <- X[, c(8)]
wheat_variety 
X <- X[, c(1,2,3,4,5,6,7)]
(X <- as.matrix(X))
```
d. 
```{r}
CovX <- cov(X)
EE <- eigen(CovX)
Evalues <- EE$values
Evec <- EE$vectors
```
So the orthogonal matrix $$ \Gamma$$ is equal to
```{r}
Gamma <- as.matrix(Evec)
Gamma
```
and the diagonal matrix $$ \Lambda $$ is
```{r}
Lambda <- diag(Evalues)
Lambda
```
Check by calculate the product of decomposition:
```{r}
Gamma%*%Lambda%*%t(Gamma)-CovX
```




Problem 2
```{r}
library(ICSNP)
data(pulmonary)
summary(pulmonary)
head(pulmonary)
```
a. 
```{r}
plot(pulmonary)
```
b. 
```{r}
help(mahalanobis)
sq_m_dist <- mahalanobis(pulmonary, center=colMeans(pulmonary), cov(pulmonary))
sq_m_dist

help(qqnorm)
qqnorm(sq_m_dist)

sq_m_dist_centered <- (sq_m_dist - mean(sq_m_dist))/sqrt(var(sq_m_dist))

test_stat = c()
for (i in 1:length(sq_m_dist_centered)) {
  test_stat[i] <- 1-pchisq(sq_m_dist_centered[i], 3)
}
test_stat
```
Based on the Normal Q-Q Plot, the line formed by data points are roughly straight, which indicates that both sets of quantiles come from the same distribution, so data may come from a multivariate normal distribution. Moreover, based on the result of Chi-squared test, all test statistics of centralized data are larger than 0.05 which means that we can not reject the null hypothesis that data come from a multiivariate normal distribution at 95% significant level. 

c. 
```{r}
help("HotellingsT2")
HotellingsT2(pulmonary)
```
P-value given by the function HotellingsT2 is 0.05123.
```{r}
n = dim(pulmonary)[1]
p = dim(pulmonary)[2]
summary(pulmonary)
S <- var(pulmonary)
X_bar <- matrix(c(mean(pulmonary[,c(1)]), mean(pulmonary[,c(2)]), mean(pulmonary[,c(3)])))
miu <- matrix(c(0,0,0)) #???
test_statistic <- n*t(X_bar-miu)%*%solve(S)%*%(X_bar-miu)
test_statistic
```
Based on Theorem 5.9, the connection between the Hotelling T2 and the F-distribution is $$ T^2(p,n) = \frac{np}{n-p+1}F_{p,n-p+1}$$
So we can calculate the corresponding F test statistic:
```{r}
F_statistic <- test_statistic/((n-1)*p/((n-1)-p+1))
F_statistic
help(pf)
1-pf(F_statistic, p, (n-1)-p+1)
```
We get the same p-value by using the connection between the Hotelling T2 and the F-distribution, both p-values are larger than 0.05 which indicate that we cannot reject the null hypothesis at 95% significant level.






















