---
title: "MAST90139 Assignment 1"
author: "Kechen Zhao 957398"
date: "29/03/2021"
output: word_document
---

Part 1
```{r}
library(readr)
domviolence <- read_csv("domviolence.csv")

# Factor each categorical variable
domviolence$reg=factor(domviolence$reg)
domviolence$age=factor(domviolence$age)
domviolence$ms=factor(domviolence$ms)
domviolence$mmo=factor(domviolence$mmo)
domviolence$smok=factor(domviolence$smok)
domviolence$alc=factor(domviolence$alc)
domviolence$falc=factor(domviolence$falc)
domviolence$educ=factor(domviolence$educ)
domviolence$dv=factor(domviolence$dv)
```
a. Fit a general model which contains all variables
```{r}
model0 <- glm(dv ~ age + ms + mmo + smok + alc + falc + educ + reg, family = binomial, domviolence)
summary(model0)
```
Use anova command to test the significance of predictors
```{r}
anova(model0, test="Chi")
```
Based on the result, we can see that predictor alc is not significant so we will drop this variable.
```{r}
model1 <- glm(dv ~ age + ms + mmo + smok + falc + educ + reg, family = binomial, domviolence)
summary(model1)
```
b. Test models' goodness of fit
```{r}
# Change age and educ as continuous variables
domviolence$age=as.integer(domviolence$age)
domviolence$educ=as.integer(domviolence$educ)
model2 <- glm(dv ~ age + ms + mmo + smok + falc + educ + reg, family = binomial, domviolence)
summary(model2)
```
Compare model2 with model1 in terms of goodness of fit using deviance: 
```{r}
anova(model1, model2, test='Chi')
```
We can see p-value is larger than 0.05, so two models are not significantly differ, we will choose model2 since it's simpler.

c. Final model construction
```{r}
# model containing all first-order interaction terms
model3 <- glm(dv ~ age + ms + mmo + smok + falc + educ + reg + (age + ms + mmo + smok + falc + educ + reg)^2, family = binomial, domviolence)
summary(model3)
```

Use step command to select the stepwise best model:
```{r}
model4 <- step(model3, trace = FALSE)
summary(model4)
```
Use anova to further drop any insignificant predictors"
```{r}
anova(model4, test='Chi')
```
We will drop predictors in smok:falc and educ:reg.
```{r}
bestmodel <- glm(formula = dv ~ age + ms + smok + falc + educ + reg + ms:falc, family = binomial, data = domviolence)
summary(bestmodel)
```

