---
title: "Asmt3"
author: "Kechen Zhao 957398"
date: "26/05/2021"
output: word_document
---

Question 1

a. The command gives a trinomial/multi-categorical logistic model with status as the nominal categorical response variable and year as the predictor. Response variable status with 3 levels follows a multinomial distribution M(371, pi) where pi is a vector that contains the probabilities of getting each category of status. 

By the coefficients given in the summary, the trinomial logit model nominal.mod treats status = normal as base level and is estimated to be:
$$log(\frac{P(mild)}{P(normal)}) = -4.2917 + 0.0836*year$$
$$log(\frac{P(severe)}{P(normal)}) = -5.0598 + 0.1093*year$$

b. 0.1093 is the coefficient of year for status = severe. Based on the estimated model written in part a, the odds of a miner's pneu-monoconiosis status is severe will increases by a factor of exp(0.1093) = 1.115497 if one more year had spent working at the coal face. 

The odds ratio of severe status vs normal status for every 10 more years spent working at the coal face is estimated to be:
$$log(\frac{P(severe)}{P(normal)}) = -5.0598 + 0.1093*year$$
$$\frac{P(severe)}{P(normal)} = e^{-5.0598 + 0.1093*year}$$
$$\frac{e^{-5.0598 + 0.1093*10}}{e^{-5.0598 + 0.1093*0}} = e^{0.1093*10} = 2.98321$$
```{r}
year = 10
odds_1b = exp(0.1093*year)
odds_1b
```
The odds ratio of severe status vs normal status for every 10 more years spent working at the coal face is estimated to be 2.98321. 

The standand error of year when status = severe is 0.0165, so the approximate 95% confidence interval is ( 2.158903, 4.122254).
```{r}
CI_log_odds = 10*(0.1093 +1.96*c(-1, 1)*0.0165)
exp(CI_log_odds)
```
c. 
Based on our model, when year = 25, 
$$log(\frac{P(mild)}{P(normal)}) = -4.2917 + 0.0836*25$$
$$\frac{P(mild)}{P(normal)} = e^{-4.2917 + 0.0836*25}$$
$$log(\frac{P(severe)}{P(normal)}) = -5.0598 + 0.1093*25$$
$$\frac{P(severe)}{P(normal)} = e^{-5.0598 + 0.1093*25}$$

```{r}
odds_mild = exp(-4.2917 + 0.0836*25)
odds_severe = exp(-5.0598 + 0.1093*25)
```
$$\frac{P(mild)}{P(normal)} + \frac{P(severe)}{P(normal)} = \frac{P(mild)+P(severe)}{P(normal)} = \frac{1-P(normal)}{P(normal)} = e^{-4.2917 + 0.0836*25} + e^{-5.0598 + 0.1093*25}$$
```{r}
# odds_mild + odds_severe 
# = prob_mild/prob_normal + prob_severe/prob_normal 
# = (prob_mild + prob_severe)/prob_normal
# = (1-prob_normal)/prob_normal
prob_normal = 1/(1+(odds_mild+odds_severe))
prob_mild = odds_mild*prob_normal
prob_severe = odds_severe*prob_normal
c(prob_normal, prob_mild, prob_severe)
prob_normal+prob_mild+prob_severe
```
The probability of status = normal is 0.82769552;
The probability of status = mild is 0.09155550;
The probability of status = severe is 0.08074898;
By summing up the above probabilities, we get 1, so the probabilities are valid. 

Question 2

a. The above command gives a cumulative logistic/proportional odds model with status as the ordinal cstegorical response variable and year as the predictor. The cdf of status' distribution is the cdf of a logistic distribution, which is equivalent to 
$$log(\frac{P(normal)}{P(mild) + P(severe)}) = 3.9558 - 0.0959*year$$
$$log(\frac{P(normal)+P(mild)}{P(severe)}) = 4.8690 - 0.0959*year$$
where all the parameter estimates above are given by the model. 

b. The odds of status moving from normal to mild/severe, or from normal/mild to severe level increases by a factor of 1/exp(-0.0959) = exp(0.0959) = 1.100649 as a coal miner spends 1 more year working at the coal face. 
```{r}
exp(0.0959)
```
The estimated odds ratio of non-normal status vs normal status for every 10 more years spent working at the coal face is:
$$log(\frac{P(normal)}{P(mild) + P(severe)}) = 3.9558 - 0.0959*year$$
$$\frac{P(normal)}{P(mild) + P(severe)} = e^{3.9558 - 0.0959*year}$$
$$\frac{P(mild) + P(severe)}{P(normal)} = \frac{1}{e^{3.9558 - 0.0959*year}}$$
$$\frac{\frac{1}{e^{3.9558 - 0.0959*10}}}{\frac{1}{e^{3.9558 - 0.0959*0}}} = \frac{e^{3.9558 - 0.0959*0}}{e^{3.9558 - 0.0959*10}} = \frac{1}{e^{-0.0959*10}} = e^{0.0959*10} = 2.609086$$
The estimated odds ratio of non-normal status vs normal status for every 10 more years spent working at the coal face is 2.609086.
The standard error for coefficient of year is 0.01194, so the approximate 95% confidence interval is (2.064682, 3.297036).
```{r}
log_CI = 10*(0.0959 + c(-1, 1)*1.96*0.01194)
exp(log_CI)
```
c. Based on the derivation before, when year = 25, we have
$$\frac{P(mild) + P(severe)}{P(normal)} = \frac{1-P(normal)}{P(normal)} = \frac{1}{e^{3.9558 - 0.0959*25}}$$
```{r}
1/(1+1/exp(3.9558 - 0.0959*25))
```
This gives P(normal) = 0.8261093.
From another equation, we have 
$$\frac{P(mild) + P(normal)}{P(severe)} = \frac{1-P(severe)}{P(severe)} = e^{4.8690 - 0.0959*25}$$
```{r}
1/(1+exp(4.8690 - 0.0959*25))
1-0.8261093-0.07788044
```
This gives P(severe) = 0.07788044. Then P(mild) = 1-P(normal)-P(severe) = 0.09601026. 

Question 3

a. Response y_it follows a Bernoulli distribution with parameter pi_it where
$$\pi_{it} = P(y_{it} = 1|age, smoke), t = 1,...,4, i = 1,...,537.$$
The mean structure of the model is
$$log(\frac{\hat{\pi_{it}}}{1-\hat{\pi_{it}}}) = -1.880-0.113*age+0.265*smoke$$
Variance is
$$\hat{Var(y_{it})} = \hat{\phi}*\pi_{it}*(1-\pi_{it}) = 0.9985*\frac{e^{-1.880-0.113*age+0.265*smoke}}{(1+e^{-1.880-0.113*age+0.265*smoke})^2}$$
The estimated correlation coefficient is (for t != s)
$$\hat{Corr}(y_{it},y_{is}) = \hat{\alpha} = 0.3543$$

b. 
```{r}
library(faraway)
data(ohio)
ohio[ohio$id == 536,]
```
The three columns represent intercept, age and smoke respectively. 
$$\begin{bmatrix} 1 & -2 & 1 \\ 1 & -1 & 1 \\ 1 & 0 & 1 \\ 1 & 1 & 1 \end{bmatrix}$$

c. The estimate of odds ratio of wheezing for a child for every one year older in age is e^(-0.113) = 0.893.
The approximate standard error of this odds ratio is
$$s.e.(\hat{odds}) = e^{\hat{\beta}}*s.e.(\hat{\beta}) = e^{-0.113}*0.044 = 0.039$$

d. The log odds ratio of wheezing for a 10-year old child with a smoking mother is
$$log.odds_{A} = -1.880-0.113*age+0.265*smoke = -1.880-0.113*1+0.265*1$$
The log odds ratio of wheezing for a 9-year old child with a non-smoking mother is
$$log.odds_{B} = -1.880-0.113*age+0.265*smoke = -1.880-0.113*0+0.265*0$$
So the odds ratio of wheezing for a 10-year old child with a smoking mother vs a 9-year old child with a non-smoking mother is
$$\frac{e^{-1.880-0.113*1+0.265*1}}{e^{-1.880-0.113*0+0.265*0}} = e^{-0.113*1+0.265*1} = 1.164$$














